Taxonomic bias can create spurious results in microbiome differential-abundance analyses unless it is properly accounted for

### 2021-08-04 abundance-measurement.Rmd
Revise manuscript (in progress)
- Begin drafting new section on measurement model
- Make minor edits and add notes to other sections
- Add refs to .bib file
- Remove summary and equations files from the rendered manuscript

@@

>>>>

## Rewrite (in progress)

Behind any quantitative measurement lies a mathematical model, either implicit or explicit, that relates the readings of our instruments and the results of our statistical analyses with the biological reality we are measuring.
This section develops a model of microbiome measurement, based on that of @mclaren2019cons, to describe how the actual absolute compositions of microbiome samples are related to estimates of relative and absolute abundance obtained by MGS.
The next section shows how these relationships impact our measurements of differential abundance.

An MGS measurement of a microbiome sample turns a collection of biological source material (cells, viruses, and extracellular DNA) into a collection of taxonomically-assigned reads (or read pairs).
For concreteness and simplicity, we suppose that all assigned reads come from cells and that the ground-truth biophysical quantity we wish to measure is the density of cells (by volume, or by wet or dry mass) belonging to various taxa, or relative abundances directly derived from these densities. 
This particular simplification reflects our own and the literature's prokaryotic bias; however, the features of MGS measurement that we explore also apply to situations (such as studies of viruses, fungi, or environmental DNA) where cell-based measurement may be inappropriate.
We next suppose that there are a set of atomic taxa, which are mutually distinct and within which we assume the properties that determine taxonomic bias are conserved.
For convenience, we identify these atomic taxa with species under a given taxonomic classification system, while acknowledging that whether taxonomic bias is conserved within commonly-accepted species has yet to be investigated in any detail.
Since DA analysis is commonly performed at a variety of taxonomic levels, we consider measurement of species and of synthetic taxa (such as a phylum or a clade in a phylogeny) that consist of multiple species.
We assume that reads assigned to a given taxon and sample really do come from that taxon and sample, but note that taxonomic mislabeling and contamination (from the reagents, lab environment, or other samples) pose further challenges to DA analysis.

- might want to note that we're mostly going to talk as if species-level taxonomic assignment is possible.

Consider the MGS measurement of a particular sample.
For a given taxon, we define the _absolute efficiency_ as the number of reads assigned to that taxon divided by the number of cells in the original sample.
The absolute efficiency depends not only on the ability of the chosen protocol to measure this taxon, but further aspects such as how much the sequencing library was diluted and how deeply the library was then sequenced.
We suppose that the absolute efficiencies of species can be written as a product of a species-specific, sample-independent factor and a species-independent, sample-specific factor.
The species-specific factors reflects the differential ability of a given protocol to measure different species (i.e. taxonomic bias), whereas the sample-specific factor captures the fact that the reads in an MGS measurement are not directly related to the absolute number or density of cells in a sample regardless of which taxa are being measured.
When we consider the _relative efficiencies_—or the ratios among absolute efficiencies—the sample-specific, taxon-independent factor cancels.
Relative efficiencies quantify how much more efficiently different taxa are measured relative to each other.
When we say efficiency unqualified, we mean efficiency relative to some fixed, arbitrary species in the denominator.

For species, the relative efficiencies are sample-independent.
In contrast, the relative efficiency of a synthetic taxon varies depending on the relative densities of its constituent species,
\begin{align}
  \text{efficiency}_T(a) = \frac{\sum_{t \in T} \text{density}_t(a) \cdot \text{efficiency}_t}{\sum_{t \in T} \text{density}_t(a)}
\end{align}
(May need to explain or derive this.)
For example, @mclaren2019cons found that for a particular shotgun-sequencing protocol, the relative efficiency was X between two Firmicute species X and Y;
hence the efficiency of Firmicutes could vary by X depending on which of these two species were dominant in a given sample.

From efficiencies, we can determine the error in relative abundances estimated by our MGS measurement.

- atomic taxa
- synthetic taxa

MGS analysis is commonly done on proportions - the density of a taxon relative to some designated taxonomic whole, such as all prokaryotes.

- give equation def of proportion
- give error in proportion, from efficiency

outline

- Relative-abundance measurements

- Absolute-abundance measurements

questions, thoughts

- should I define absolute efficiency in terms of cells or density?
- need to revise the story of the first paragraphs above.

## Current working version

@@

**Update:** Actually, the two methods are equivalent when the total density and proportion are estimated using the same reads.
Let $B$ be all bacteria and $H$ be the host.
Consider using WGS reads to estimate the absolute abundance of bacterial taxon $i$,
\begin{align}
  \hat{\text{density}}_{i}(s) 
  &= 
  \underbrace{\frac{\text{reads}_{B}}{\text{reads}_{H}}}_\text{est'd bact'l density} \cdot 
  \underbrace{\frac{\text{reads}_{i}}{\text{reads}_{B}}}_\text{est'd prop'n}
  = \frac{\text{reads}_{i}}{\text{reads}_{H}}
\end{align}
So the problem/difference arises when we the proportion is estimated either by a different technology (e.g. when density is estimated via WGS and proportion is estimated by 16S sequencing) or the two estimates going into the above are not just ratios of reads.
For example, ratios of bacterial to host reads could be used for the density estimate, while the proportion is estimated by Metaphlan.

### 2021-08-04 differential-abundance.Rmd
Revise manuscript (in progress)
- Begin drafting new section on measurement model
- Make minor edits and add notes to other sections
- Add refs to .bib file
- Remove summary and equations files from the rendered manuscript

@@

<!-- P: error in FC ratios and ratio-based AAs -->
(In contrast to proportion-based estimates) Becuase consistent taxonomic bias creates a consistent error in ratio-based estimates, the error completely cancels, leaving accurate fold changes.
From \@ref(eq:ratio-err), the estimated fold change in the ratio of taxon $i$ to taxon $j$ from sample $s$ to sample $t$ is
\begin{align}
  (\#eq:ratio-fc-err)
  \underbrace{ \frac{{\text{reads}_{i}(t)}/{\text{reads}_{j}(t)}} {{\text{reads}_{i}(s)}/{\text{reads}_{j}(s)}} }_\text{estimate}
  &= 
  \frac{{\text{density}_{i}(t)}/{\text{density}_{j}(t)}} {{\text{density}_{i}(s)}/{\text{density}_{j}(s)}}
  \cdot
  \frac{\cancel{\text{efficiency}_{i}(t)}/\cancel{\text{efficiency}_{j}(t)}} {\cancel{\text{efficiency}_{i}(s)}/\cancel{\text{efficiency}_{j}(s)}}
  \\[0.5ex]&= 
  \underbrace{ \frac{{\text{density}_{i}(t)}/{\text{density}_{j}(t)}} {{\text{density}_{i}(s)}/{\text{density}_{j}(s)}} }_\text{actual}
  .
\end{align}
(Describe FCs in Figure 2 and Figure 3 density e.g., seeing the correct FCs.)

@@

Interestingly, taxonomic bias may actually make qPCR a better method for absolute quantification than cell counting for 16S rRNA gene sequencing experiments.
16S rRNA gene qPCR measures the concentration of 16S rRNA gene copies in the extracted DNA and is therefore necessarily affected by three large sources of bias in 16S rRNA gene experiments: extraction, amplification, and copy-number variation.
Yet these biases are shared by the sequencing measurement.
Though they make qPCR measurements a bad proxy for total cell density^[See note from KL below], when used for absolute differential abundance inference the shared bias in the qPCR and sequencing measurements can offset each other, leading to estimated fold changes that are relatively unaffected by taxonomic bias.
(Jian, Salonen, and Korpela 2021 suggests some of these ideas and APPENDIX gives a mathematical justification.)
If common primers are used by qPCR and 16S rRNA gene sequencing, the bias due to primer mismatches and other sources of variation in amplification could even be accounted for in this manner.^[See discussion below]
Yet ideally we would still be able to estimate and correct remaining error due to unshared bias or to mechanisms, such as saturation during DNA extraction, that could break the assumed proportionality between qPCR measurements and total abundance.
In this case, ...

## Notes

<details><summary style="font-weight:bold;font-size:120%">Details: Taxonomic bias and differential-abundance estimates</summary>
::: {#details:differential .rmdnote}

#### Proportion-based inference {-}

#### Ratio-based inference {-}

:::
</details>

>>>>

_The current text generally assumes that taxa are atomic._

@@

<!-- P: error in FC ratios and ratio-based AAs -->
(In contrast to proportion-based estimates) Because consistent taxonomic bias creates a consistent error in ratio-based estimates, the error completely cancels, leaving accurate fold changes.
From \@ref(eq:ratio-err), the estimated fold change in the ratio of taxon $i$ to taxon $j$ from sample $s$ to sample $t$ is
\begin{align}
  (\#eq:ratio-fc-err)
  \underbrace{ \frac{{\text{reads}_{i}(t)}/{\text{reads}_{j}(t)}} {{\text{reads}_{i}(s)}/{\text{reads}_{j}(s)}} }_\text{estimate}
  &= 
  \frac{{\text{density}_{i}(t)}/{\text{density}_{j}(t)}} {{\text{density}_{i}(s)}/{\text{density}_{j}(s)}}
  \cdot
  \frac{\cancel{\text{efficiency}_{i}}/\cancel{\text{efficiency}_{j}}} {\cancel{\text{efficiency}_{i}}/\cancel{\text{efficiency}_{j}}}
  \\[0.5ex]&= 
  \underbrace{ \frac{{\text{density}_{i}(t)}/{\text{density}_{j}(t)}} {{\text{density}_{i}(s)}/{\text{density}_{j}(s)}} }_\text{actual}
  .
\end{align}
(Describe FCs in Figure 2 and Figure 3 density e.g., seeing the correct FCs.)
(_Need to make plain in the equation that this is the estimated fold change in the density of taxon i; perhaps have notation for this._)

### Synthetic or higher-level taxa

_Some of these concepts would best be introduced in the previous secton on abundances.
Alternatively, we might want to separate them into a box or appendix on multiplicative aggregation._

<!-- P: Synthetic taxa intro -->

Frequently, relative- and absolute-abundance estimation are used to infer fold changes in higher-level taxa.

- Visual or statistical estimation of changes in phylum abundance is a ubiquitous form of analysis in bacterial microbiome studies.
- @chng2020meta - use fold changes in the ratio of bacterial to host reads to estimate the change in bacterial biomass during mouse antibiotics experiment
- @rao2021mult - use a spike-in to infer changes in total domain abundance for each of archaea, bacteria, and fungi.
- @karasov2020ther - ratio of microbial to host reads to infer changes in microbial biomass

We know that efficiencies vary widely within domains and also has been observed to vary within phyla and even within species.
What are the implications of this within-taxon variation for infered fold changes?

<!-- P: Synthetic taxa results -->

Unfortunately, the bias invariance of ratio-based inference of species need not apply to ratios involving higher-level taxa.
The estimated fold change in the proportion, ratio, or density of a synthetic taxon $T$ has additional multiplicative error equal to its fold change in efficiency.
This error can be substantial when there is a substantial turnover from dominance by high- to low-efficiency species (or vice versa).
(todo: ref relevant equation from previous section)

An alternative approach to creating synthetic quantitative features was develoepd in the Compositional Data Analysis field that retains the so-called perturbation invariance property of atomic features.
These "balances" are constructed by multiplying, rather than adding, the abundances of the atomic elements (equivalent to adding the abundances after log transformation instead of before) (@vandenboogaart2013anal).
The changes in balances retain bias invariance (@mclaren2019cons Appendix 1, and REF Appendix).
Balances have been used to construct candidate disease biomarkers (@riverapinto2018bala, @quinn2020inte), develop phylogenetic community distance measures (@silverman2017aphy), and look for clades associated with traits (@washburne2017phyl). 
But they face from conceptual and practice issues relating to the sparsity of microbiome data. 
Perhaps more important, products of species abundances do not obviously map onto the ecological notion of abundance and hence do not seem to fully substitute for attempting to infer the changes in taxon abundance as conceived in the usual sense of additive aggregation.
A benefit of products, however, is that they are also robust to the fact that we may not know the appropriate ecological conversion rate for different species; for example, cells of one clade might be twice as impactful per cell on a host or  ecosystem as cells from another clade.
Grappling with these issues remains an important goal for future theoretical and empirical research.

@@

Interestingly, taxonomic bias may actually make qPCR a better method for absolute quantification than cell counting for 16S rRNA gene sequencing experiments.
16S rRNA gene qPCR measures the concentration of 16S rRNA gene copies in the extracted DNA and is therefore necessarily affected by three large sources of bias in 16S rRNA gene experiments: extraction, amplification, and copy-number variation.
Yet these biases are shared by the sequencing measurement.
Though they make qPCR measurements a bad proxy for total cell density, when used for absolute differential abundance inference the shared bias in the qPCR and sequencing measurements can offset each other, leading to estimated fold changes that are relatively unaffected by taxonomic bias.
(Jian, Salonen, and Korpela 2021 suggests some of these ideas and APPENDIX gives a mathematical justification.)
If common primers are used by qPCR and 16S rRNA gene sequencing, the bias due to primer mismatches and other sources of variation in amplification could even be accounted for in this manner.
Yet ideally we would still be able to estimate and correct remaining error due to unshared bias or to mechanisms, such as saturation during DNA extraction, that could break the assumed proportionality between qPCR measurements and total abundance.
In this case, ...

### 2021-08-04 solutions.Rmd
Revise manuscript (in progress)
- Begin drafting new section on measurement model
- Make minor edits and add notes to other sections
- Add refs to .bib file
- Remove summary and equations files from the rendered manuscript

@@

_Except in synthetic community experiments, we are always likely to have taxa not in our controls. We can attempt to use phylogenetic prediction or other modeling of efficiencies to predict the efficiencies of other taxa; however there will be signifciant undertainty. Therefore our control measurements may ideally serve to inform a bias-sensitivity analysis of the sort described below._

For revision:

- Discuss the vaginal mb and leopold examples

@@

targeted measurements of a small number of taxa

Because targeted absolute-abundance measurements are expected to provide (relatively) accurate  estimates of fold changes, they can be used to validate and even correct the DA results derived from a proportion-based absolute DA analysis.
We illustrate the basic idea using the @lloyd2020evid experiment described above.
Absolute abundances in this study were estimated by multiplying MGS proportions by total abundance measured by cell counting; for comparison, qPCR was used to measure specific microbial taxa.^[see note]
Because the error in fold changes or a regression from the MGS measurements is taxon-independent, the difference between the two methods for these reference taxa informs us about the difference for all taxa.
@lloyd2020evid found close agreement for the specific taxa also measured by qPCR (Table 1), suggesting that variation in sample mean efficiency did not significantly distort their results.
In principle, a joint statistical analysis of all measurements would allow inferring variation in the sample mean efficiency across samples and obtaining calibrated fold change and regression estimates for all taxa [APPENDIX].

### Notes for revision

@@

> Might ratio-based analysis also be used to overcome the effects of bias in differential absolute abundance? In fact, some methods for determining absolute abundance are based on ratios among taxa in the MGS measurement instead of proportions. Rather than using a measurement of total abundance, these methods require determining the abundance of one or more reference taxa.

This needs to be revised to make clear the novelty of using non-spiked taxa as a reference; could help to replace "In fact, some methods" with clearer language.
Also make clear that some studies using reference taxa are using proportion-based methods.
Also, cite the appendix that gives an overview of AA methods, and explicitly state distinction between proportion-based and ratio-based methods.

Could help to change to:

- Ratio-based analysis can also be used to overcome the effects of bias in differential absolute abundance.
- /Make clear that the above ratio equation applies for relative and absolute observed or actual abudances (since totals cancel)/
- Suppose we knew the true absolute abundance of a reference taxon $r$; then we might estimate the absolute abundance of taxon $i$ from
  + Give the estimation equation
  + Explain the effect of taxonomic bias - get correct FCs
- It remains to determine the AA of one or more reference taxa.
  + Some existing AA methods involve using spike ins of extraneous taxa
    - discuss spike-in approach
    - often stated that bias a problem for spike-ins; however we see that it only affects the baseline and not the fold changes
      + this first part could be shunted to a discussion
    - most/all papers on spike-ins suppose the spike-in must be added in a known fixed amount.
    - however, the above argument indicates that it isn't required that the reference taxon is extraneous or has a fixed abundance; what is important is that we know the correct abundance up to a fixed multiplicative error
    - hence it can be a native taxon we've measured using a targeted method; this method can be biased as long as it has a fixed multiplicative error (unlike MGS bias.)

Need to consider if this is too detailed for the main text.
Might want to make this section of the appendix particularly clear if can't put this in the main text.

Key equations, to show side-by-side: Proportion-based and ratio-based AA inference.

@@

- from gdoc: this distinction of ratios vs proportions needs to be spelled out more explicitly; could even use my seminar figure (or equations) to help
- later thoughts: This distinction is so central for how I understand and explain the key ideas, should refer to it early and often

@@

>>>>

_This section needs to be revised to account for the expanded earlier theory, especially relating to targeted measurements and ratio-based methods._

@@

- Except in synthetic community experiments, we are always likely to have taxa not in our controls. We can attempt to use phylogenetic prediction or other modeling of efficiencies to predict the efficiencies of other taxa; however there will be signifciant undertainty. Therefore our control measurements may ideally serve to inform a bias-sensitivity analysis of the sort described below.
- Should mention the two extant cases where mocks of relevant taxa have been created: The vaginal-HMP / MOMS-PI study (@brooks2015thet and @fettweis2019thev) and the @leopold2020host plant-fungal experiment. Note that even in these cases, calibration hasn't been much used, partly because of a lack of workflows or understanding of when and why it is necessary. Ideally we can give an illustration; this section can be a good use of the @leopold2020host change-over-time result, though I'm not sure whether to discuss that here or in the 'real-world' section.

@@

Because targeted absolute-abundance measurements are expected to provide (relatively) accurate estimates of fold changes, they can be used to validate and even correct the DA results derived from a proportion-based absolute DA analysis.
We illustrate the basic idea using the @lloyd2020evid experiment described above.
Absolute abundances in this study were estimated by multiplying MGS proportions by total abundance measured by cell counting; for comparison, qPCR was used to measure specific microbial taxa.
Because the error in fold changes or a regression from the MGS measurements is taxon-independent, the difference between the two methods for these reference taxa informs us about the difference for all taxa.
@lloyd2020evid found close agreement for the specific taxa also measured by qPCR (Table 1), suggesting that variation in sample mean efficiency did not significantly distort their results.
In principle, a joint statistical analysis of all measurements would allow inferring variation in the sample mean efficiency across samples and obtaining calibrated fold change and regression estimates for all taxa [APPENDIX].

@@

@@

@@

- Aim to illustrate with a simulated or real example
